GWAS datasets for FDR benchmarking

### 2017-09-20 datasets/GWAS/GWAS.Rmd
Add benchmarking evaluation for GWAS data

@@

Here we download the GWAS BMI dataset analyzed in the Boca-Leek paper.

@@

```{r, GWAS1-verify-contents}
library(data.table)
bmi <- fread("BMI.SNPadjSMK.CombinedSexes.AllAncestry.txt", header=TRUE)
dim(bmi)
head(bmi)
```

@@

```{r, GWAS1-cleanup}
rm(bmi)
```

@@

@@

```{r, GWAS2-verify-contents}
# read in meta-data
library(readxl)
meta <- read_xls("gabriel_results_description.xls")
head(meta)
dim(meta)

@@

# remove study-specific columns
library(dplyr)
empty <- which(!is.na(meta$description))
asthma <- asthma %>%
            select(empty)
meta <- meta %>% 
            slice(empty)
head(asthma)
```

@@

- `P_fix` and `P_ran`:p-values for fixed and random effect models
- `theta_fix` and `theta_fix`:effect size for fixed and random effect models
- `se_theta_fix` and `se_theta_fix`:standard error for fixed and random effect models
- additional covariates: 
    - `no_studies_meta`: Number of studies included in the meta-analysis with this SNP
    - `freq_all_1_max` and `freq_all_1_min`: Maximum and Minimum Allele Frequency of Allele 1 across samples

for 567,589 SNPs.

```{r, GWAS2-cleanup}
rm(asthma)
```

@@

```{r, GWAS3-cleanup}
rm(trig)
```

>>>>

We'll also run the benchmark methods on each dataset and report the number of 
discoveries for each one.

### Workspace setup

To run the benchmark methods on the GWAS datasets, we'll use some of the helper 
and wrapper functions from the simulated data analysis. We also create and source an analgous version of the `sim_runner` function to apply to the real data.

```{r, wkspace-setup}
# Note that swfdr package depends on R version 3.4 or higher
setwd("/n/irizarryfs01_backed_up/kkorthauer/benchmark-fdr/")
source("simulation_studies/R/simulation-helpers.R")
source("datasets/R/evaluation-helpers.R")
setwd("./datasets/GWAS")

# wrangling & plotting tools
library(data.table)
library(readxl)
library(dplyr)
library(ggplot2)

# benchmark methods
library(IHW)
library(ashr)
library(qvalue)
library(swfdr)
library(fdrtool)
```

@@

Here we download and analyze the GWAS BMI dataset analyzed in the Boca-Leek paper.

#### Data download

@@

```{r, GWAS1-verify-contents}
bmi <- fread("BMI.SNPadjSMK.CombinedSexes.AllAncestry.txt", header=TRUE)
dim(bmi)
head(bmi)
```

@@

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of subjects with that SNP
measured). **Still to-do: consider multiple covariates for methods that allow this.**

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N)
```

Now, we'll run the `run_benchmarks` helper function on the `bmi` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods (since there are more than 2.5 million tests)
and best run on a machine with lots of memory.

```{r, GWAS1-run}
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
if (!file.exists("bmi_results.RData")){
  bmi_results <- run_benchmarks(dat=bmi, alphas=cutoffs)
  save(bmi_results, file="bmi_results.RData")
}else{
  load("bmi_results.RData")
}

# plot
ggplot(bmi, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("BMI GWAS Metanalysis with Sample Size Covariate")
```

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS1-cleanup}
rm(bmi)
rm(asthma_results)
```

@@

#### Data Download

@@

```{r, GWAS2-verify-contents}
# read in meta-data
meta <- read_xls("gabriel_results_description.xls")
head(meta)
dim(meta)

@@

# remove study-specific columns
empty <- which(!is.na(meta$description))
asthma <- asthma %>%
            select(empty)
meta <- meta %>% 
            slice(empty)

# remove SNPs with missing effect size/pvalues
asthma <- asthma %>% 
            filter(!is.na(P_fix) & !is.na(theta_fix))

dim(asthma)
head(asthma)
```

@@

- `P_fix` and `P_ran`:p-values for fixed and random effect models
- `theta_fix` and `theta_ran`:effect size for fixed and random effect models
- `se_theta_fix` and `se_theta_ran`:standard error for fixed and random effect models
- additional covariates: 
    - `no_studies_meta`: Number of studies included in the meta-analysis with this SNP
    - `freq_all_1_max` and `freq_all_1_min`: Maximum and Minimum Allele Frequency of Allele 1 across samples

for 546,182 SNPs.

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of studies with that SNP
measured). We'll also focus on the fixed effects tests first.
**Still to-do: consider multiple covariates for methods that allow this.**

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta)
```

Now, we'll run the `run_benchmarks` helper function on the `asthma` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods
(since there are more than 500 thousand tests)
and best run on a machine with lots of memory.

```{r, GWAS2-run}
# use same cutoffs as for the bmi study (defined above)
if (!file.exists("asthma_results.RData")){
  asthma_results <- run_benchmarks(dat=asthma, alphas=cutoffs)
  save(asthma_results, file="asthma_results.RData")
}else{
  load("asthma_results.RData")
}

# plot
ggplot(asthma_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate")
```

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS2-cleanup}
rm(asthma)
rm(asthma_results)
```

@@

#### Running benchmark methods

First, we'll standardize the data frame to use the wrapper function. For now, we'll
assume that there is only a single 'independent covariate' for those methods that 
utilize this, and we'll use the sample size (number of individuals with that SNP
measured). 
**Still to-do: consider multiple covariates for methods that allow this.**

```{r, GWAS3-dat}
trig <- trig %>% 
        rename(pval = P_value,
               SE = SE,
               effect_size = beta,
               ind_covariate = N)
```

Now, we'll run the `run_benchmarks` helper function on the `trig` data frame for
several alpha values and plot the results. This is fairly computationally intensive 
for some of the benchmark methods
(since there are more than 500 thousand tests)
and best run on a machine with lots of memory.

```{r, GWAS2-run}
# use same cutoffs as for the bmi study (defined above)
if (!file.exists("triglyceride_results.RData")){
  trig_results <- run_benchmarks(dat=trig, alphas=cutoffs)
  save(trig_results, file="triglyceride_results.RData")
}else{
  load("triglyceride_results.RData")
}

# plot
ggplot(trig_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate")
```

And we'll clean up the workspace before moving on to the next dataset.

```{r, GWAS3-cleanup}
rm(trig)
rm(trig_results)
```
